---
title: "Compositional DA Method Comparison: HRSSL vs NN"
author: "Brad"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
geometry: margin=2.2cm
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5,
  fig.pos = "H", out.extra = "",
  dev = "cairo_pdf"
)
options(knitr.kable.NA = "")
```

```{r libs}
# Bioconductor / S4Vectors-touching packages FIRST so tidyverse can mask cleanly.
# New packages for this report (install once):
#   BiocManager::install("maaslin3")          # MaAsLin 3
#   BiocManager::install("ANCOMBC")           # ANCOM-BC2 (ancombc2()), pulls in
#                                              # TreeSummarizedExperiment
#   install.packages("MicrobiomeStat")        # provides linda()
#   install.packages("callr")                 # run_maaslin3() uses a clean
#                                              # subprocess (see Methods)
suppressPackageStartupMessages({
  library(MicrobiomeStat)   # linda()
  library(maaslin3)         # maaslin3()
  library(ANCOMBC)          # ancombc2()
  library(TreeSummarizedExperiment)
  library(ggplot2)
  library(ggrepel)
  library(kableExtra)
  # tidyverse / dplyr LAST
  library(tidyverse)
})

# Explicit dplyr namespacing convention - use dplyr:: in code chunks.
theme_set(theme_bw(base_size = 10) +
          theme(panel.grid.minor = element_blank(),
                strip.background = element_rect(fill = "grey95", colour = NA)))

group_colours <- c("HRSSL" = "#C0392B", "NN" = "#2E86AB")

q_cut <- 0.25   # FDR threshold; preserves sensitivity at n = 12/group
```

```{r paths}
# ---- EDIT THESE PATHS --------------------------------------------------------
proc_dir <- "humann3_processed"

path_cpm_unstrat_file <- file.path(proc_dir, "path_split_cpm",
                                   "all_pathabundance_cpm_named_unstratified.tsv")
ko_cpm_unstrat_file   <- file.path(proc_dir, "ko_split_cpm",
                                   "all_ko_cpm_named_unstratified.tsv")

# Un-normalised RPK versions, used by ANCOM-BC2 only (see Methods). Confirm
# these match the actual RPK output names from 01_humann3_postprocess.sh.
path_rpk_unstrat_file <- file.path(proc_dir, "path_split_rpk",
                                   "all_pathabundance_unstratified.tsv")
ko_rpk_unstrat_file   <- file.path(proc_dir, "ko_split_rpk",
                                   "all_ko_unstratified_named.tsv")

metadata_file <- "nnandhrsslmetadata.csv"   # SampleID, Group (HRSSL / NN)

out_dir <- "humann3_da_comparison_out"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
# -----------------------------------------------------------------------------
```

# Overview

This report compares three compositional differential-abundance methods on the
same HRSSL vs NN functional profiles, on the MetaCyc pathway and KEGG ortholog
(KO) layers. The earlier richness-covariate approach has been retired in favour
of methods that handle the closure problem inside the model rather than through
an added fixed effect:

\begin{description}
  \item[\textbf{LinDA}] a linear model on log-ratio transformed data that then
        corrects the compositional bias term using the mode of the coefficient
        distribution, recovering the true reference frame.
  \item[\textbf{MaAsLin 3}] separates prevalence from abundance and corrects
        abundance coefficients by comparing each against the median coefficient
        for the variable across features (\texttt{median\_comparison\_abundance}).
  \item[\textbf{ANCOM-BC2}] a log-linear model that estimates a sample-specific
        sampling-fraction (bias) term via iterative bias correction and
        subtracts it before testing, then applies a Holm/BH-adjusted Wald test
        to the bias-corrected coefficients. It is given the un-normalised RPK
        abundances, since the sampling-fraction step is meant to model the
        depth variation that CPM normalisation has already removed.
\end{description}

All three are fed an identical feature set: \texttt{drop\_unmapped()} applied,
then a single prevalence filter (detected in at least 25\% of samples in either
group). LinDA and MaAsLin 3 receive these features as CPM; ANCOM-BC2 receives
the same features as un-normalised RPK, because its bias correction is designed
to estimate the per-sample sampling fraction from depth-carrying abundances and
CPM normalisation removes exactly that signal. The internal prevalence filters
of each method are set to zero so none re-filter. Effect-size magnitudes are not
on the same scale across methods, so agreement is judged by significance
overlap, direction, and rank rather than by matching coefficients. Significance
throughout is taken at \(q < `r q_cut`\).

The three methods recover the reference frame by different routes (LinDA via the
mode of the coefficient distribution, MaAsLin 3 via the median comparison,
ANCOM-BC2 via the estimated sampling fraction), so close agreement among them is
the signal worth trusting. The consensus set, features called by all three, is
the defensible core; features called by only one method warrant the most
caution.

# Data ingestion

```{r read-helper}
# HUMAnN tables: one feature column then sample columns. Read robustly, strip
# "# " comment from header, coerce to matrix with rownames = features.
read_humann <- function(path) {
  tab <- read.delim(path, sep = "\t", header = TRUE,
                    check.names = FALSE, comment.char = "")
  names(tab)[1] <- "Feature"
  tab$Feature <- sub(":\\s*NO_NAME\\s*$", "", tab$Feature)
  feat_col <- tab$Feature
  mat <- as.matrix(tab[, -1, drop = FALSE])
  rownames(mat) <- feat_col
  mat
}

drop_unmapped <- function(mat) {
  keep <- !rownames(mat) %in% c("UNMAPPED", "UNINTEGRATED", "UNGROUPED",
                                "UNMAPPED|unclassified",
                                "UNINTEGRATED|unclassified",
                                "UNGROUPED|unclassified")
  mat[keep, , drop = FALSE]
}
```

```{r read-data}
path_cpm <- drop_unmapped(read_humann(path_cpm_unstrat_file))
ko_cpm   <- drop_unmapped(read_humann(ko_cpm_unstrat_file))

# RPK (un-normalised) versions for ANCOM-BC2.
path_rpk <- drop_unmapped(read_humann(path_rpk_unstrat_file))
ko_rpk   <- drop_unmapped(read_humann(ko_rpk_unstrat_file))

meta <- read.csv(metadata_file, stringsAsFactors = FALSE)
stopifnot(all(c("SampleID", "Group") %in% names(meta)))
meta$SampleID <- trimws(as.character(meta$SampleID))
meta$Group    <- toupper(trimws(as.character(meta$Group)))
meta$Group    <- factor(meta$Group, levels = c("NN", "HRSSL"))
meta <- meta[!is.na(meta$Group) & nzchar(meta$SampleID), , drop = FALSE]
rownames(meta) <- meta$SampleID

# Restrict to samples present in metadata and every input matrix.
common <- Reduce(intersect, list(meta$SampleID,
                                 colnames(path_cpm), colnames(ko_cpm),
                                 colnames(path_rpk), colnames(ko_rpk)))
stopifnot(length(common) >= 4)
samples <- common
meta    <- meta[samples, , drop = FALSE]

# Match features across the two normalisations of each layer, and restrict both
# to the common samples, so CPM and RPK share an identical feature set and order.
path_feats <- intersect(rownames(path_cpm), rownames(path_rpk))
ko_feats   <- intersect(rownames(ko_cpm),   rownames(ko_rpk))
path_cpm <- path_cpm[path_feats, samples, drop = FALSE]
path_rpk <- path_rpk[path_feats, samples, drop = FALSE]
ko_cpm   <- ko_cpm[ko_feats, samples, drop = FALSE]
ko_rpk   <- ko_rpk[ko_feats, samples, drop = FALSE]
```

# Prevalence filtering

The prevalence and variance filters are computed on the RPK layer, which is the
input ANCOM-BC2 receives and the only normalisation-sensitive method here.
Detection is unchanged by per-sample normalisation, so the prevalence step is
identical on CPM. Features detected (>0) in at least 25% of samples *in either
group* are kept, matching the existing pipeline. Features whose detected
(non-zero) values are constant within either group are then dropped: ANCOM-BC2
estimates each taxon's variance from its observed values, so such a feature
leaves zero residual variance once the group means are fitted and makes the
iterative bias-correction step fail outright (it errors with "Zero variances
have been detected"). A raw within-group variance check does not catch these,
because a feature that mixes zeros with a single repeated non-zero value still
has non-zero variance overall. The surviving feature set is applied unchanged to
the CPM matrices used by the other two methods, so all three see an identical
feature set.

```{r prevalence}
prev_filter <- function(mat, groups, min_prev = 0.25) {
  grp <- split(colnames(mat), groups)
  keep <- vapply(rownames(mat), function(f) {
    any(vapply(grp, function(s) mean(mat[f, s] > 0) >= min_prev, logical(1)))
  }, logical(1))
  mat[keep, , drop = FALSE]
}

# Drop features whose detected (non-zero) values are constant within either
# group. ANCOM-BC2 estimates each taxon's variance from its observed
# (non-zero) log-abundances, so a feature can pass a raw variance check
# (non-zero only because it mixes zeros with a single repeated value) and still
# leave zero residual variance once the group means are fitted, which is what
# triggers "Zero variances have been detected for the following taxa ...".
# Requiring at least two distinct detected values in each group removes that
# failure mode and keeps the feature set identical across all three methods.
zero_var_filter <- function(mat, groups) {
  grp <- split(colnames(mat), groups)
  keep <- vapply(rownames(mat), function(f) {
    all(vapply(grp, function(s) {
      present <- mat[f, s]
      present <- present[present > 0]
      length(present) >= 2L && stats::var(present) > 0
    }, logical(1)))
  }, logical(1))
  mat[keep, , drop = FALSE]
}

# The prevalence and variance filters are computed on the RPK layer, which is
# what ANCOM-BC2 receives and the only normalisation-sensitive method here.
# Detection is unchanged by per-sample scaling, so the prevalence step is
# identical on CPM. The surviving feature set is then applied unchanged to the
# CPM matrices used by the other two methods.
path_rpk_p <- prev_filter(path_rpk, meta$Group)
ko_rpk_p   <- prev_filter(ko_rpk,   meta$Group)

path_rpk_f <- zero_var_filter(path_rpk_p, meta$Group)
ko_rpk_f   <- zero_var_filter(ko_rpk_p,   meta$Group)

path_cpm_f <- path_cpm[rownames(path_rpk_f), , drop = FALSE]
ko_cpm_f   <- ko_cpm[rownames(ko_rpk_f),     , drop = FALSE]

# Name maps to undo MaAsLin's make.names() sanitisation back to the original
# "PWY-ID: name" / "K-ID: name" strings, so all three methods share join keys.
path_name_map <- stats::setNames(rownames(path_cpm_f), make.names(rownames(path_cpm_f)))
ko_name_map   <- stats::setNames(rownames(ko_cpm_f),   make.names(rownames(ko_cpm_f)))

data.frame(
  Layer  = c("Pathways", "KOs"),
  Before = c(nrow(path_rpk), nrow(ko_rpk)),
  `After prevalence` = c(nrow(path_rpk_p), nrow(ko_rpk_p)),
  `After variance`   = c(nrow(path_rpk_f), nrow(ko_rpk_f)),
  Samples = c(ncol(path_rpk_f), ncol(ko_rpk_f)),
  check.names = FALSE
) |>
  kbl(booktabs = TRUE, caption = "Features before and after >=25% prevalence in either group, then after dropping features with constant detected values within either group. Counts are from the RPK layer, which determines the shared feature set.") |>
  kable_styling(latex_options = c("hold_position"))
```

# Methods

```{r runner-helpers}
# Each runner returns a tidy tibble: feature, effect (method-native, signed so
# that positive = higher in HRSSL), pvalue, qvalue, method. Feature keys are the
# original ID strings in every case, so the three are directly joinable.

# --- LinDA: bias-corrected CLR linear model; effect = log2 fold change -------
run_linda <- function(mat, meta, ref = "NN") {
  md <- meta[colnames(mat), , drop = FALSE]
  md$Group <- stats::relevel(factor(md$Group), ref = ref)
  non_ref  <- setdiff(levels(md$Group), ref)[1]
  set.seed(1)
  fit <- MicrobiomeStat::linda(
    feature.dat       = mat,                 # features x samples
    meta.dat          = md,
    formula           = "~Group",
    feature.dat.type  = "proportion",        # CPM; CLR step is scale-invariant
    prev.filter       = 0,
    mean.abund.filter = 0,
    max.abund.filter  = 0,
    is.winsor         = TRUE,
    outlier.pct       = 0.03,
    p.adj.method      = "BH",
    verbose           = FALSE
  )
  res <- fit$output[[paste0("Group", non_ref)]]
  tibble::tibble(
    feature = rownames(res), effect = res$log2FoldChange,
    pvalue  = res$pvalue, qvalue = res$padj, method = "LinDA"
  )
}

# --- MaAsLin 3: median-comparison compositional correction (its default) -----
# Run in a fresh R session via callr. maaslin3() registers `logging` package
# handlers/connections for its "maaslin3.log" file and does not fully tear
# them down on exit; a second in-session call (pathways already ran) then
# errors with "cannot open the connection" when it tries to write its log.
# A clean subprocess sidesteps that global-state leakage entirely.
run_maaslin3 <- function(mat, meta, out_sub, name_map, ref = "NN") {
  od <- file.path(out_dir, out_sub); unlink(od, recursive = TRUE)
  dir.create(od, recursive = TRUE, showWarnings = FALSE)
  md <- meta[colnames(mat), , drop = FALSE]
  md$Group <- factor(md$Group)
  non_ref  <- setdiff(levels(md$Group), ref)[1]

  res <- callr::r(
    function(mat, md, od, ref) {
      set.seed(1)
      fit <- maaslin3::maaslin3(
        input_data     = as.data.frame(t(mat)),  # samples x features
        input_metadata = md,
        output         = od,
        formula        = "~Group",
        reference      = paste0("Group,", ref),
        normalization  = "TSS",                   # recommended for relative data
        transform      = "LOG",
        median_comparison_abundance = TRUE,        # the compositional correction
        correction     = "BH",
        plot_summary_plot = FALSE,
        plot_associations = FALSE,
        cores          = 1
      )
      fit$fit_data_abundance$results
    },
    args = list(mat = mat, md = md, od = od, ref = ref)
  )

  res <- res[res$metadata == "Group" & res$value == non_ref, , drop = FALSE]
  tibble::tibble(
    feature = dplyr::coalesce(unname(name_map[res$feature]), res$feature),
    effect  = res$coef, pvalue = res$pval_individual,
    qvalue  = res$qval_individual, method = "MaAsLin3"
  )
}

# --- ANCOM-BC2: iterative bias correction; effect = bias-corrected log fold change ---
run_ancombc2 <- function(mat, meta, name_map, ref = "NN") {
  md <- meta[colnames(mat), , drop = FALSE]
  md$Group <- stats::relevel(factor(md$Group), ref = ref)
  non_ref  <- setdiff(levels(md$Group), ref)[1]
  tse <- TreeSummarizedExperiment::TreeSummarizedExperiment(
    assays  = list(counts = mat),   # un-normalised RPK; ANCOM-BC2 estimates the
    colData = md                    # per-sample sampling fraction itself, so it
  )                                 # expects depth-carrying input, not CPM
  set.seed(1)
  out <- ANCOMBC::ancombc2(
    data         = tse,
    assay_name   = "counts",
    fix_formula  = "Group",
    p_adj_method = "BH",
    prv_cut      = 0,
    lib_cut      = 0,
    group        = "Group",
    struc_zero   = FALSE,
    neg_lb       = FALSE,
    alpha        = q_cut,
    pseudo_sens  = FALSE,
    n_cl         = 1,
    verbose      = FALSE
  )
  res     <- out$res
  lfc_col <- paste0("lfc_Group", non_ref)
  p_col   <- paste0("p_Group", non_ref)
  q_col   <- paste0("q_Group", non_ref)
  tibble::tibble(
    feature = dplyr::coalesce(unname(name_map[res$taxon]), res$taxon),
    effect  = res[[lfc_col]], pvalue = res[[p_col]],
    qvalue  = res[[q_col]], method = "ANCOM-BC2"
  )
}
```

```{r summary-helpers}
method_levels <- c("LinDA", "MaAsLin3", "ANCOM-BC2")

# q-value formatter (fixed notation for the typical BH range, scientific for
# very small values), matching the existing report.
fmt_q <- function(q) ifelse(is.na(q), "",
  ifelse(q >= 1e-3, sprintf("%.3f", q), formatC(q, format = "e", digits = 2)))
dir_token <- function(x) dplyr::case_when(is.na(x) ~ "", x > 0 ~ "Up",
                                          x < 0 ~ "Dn", TRUE ~ "")

add_flags <- function(long) long |>
  dplyr::mutate(
    method = factor(method, levels = method_levels),
    sig    = !is.na(qvalue) & qvalue < q_cut,
    Dir    = dir_token(effect)
  )

# Pairwise concordance: features called by both, sign agreement among those,
# and Spearman of effect over the shared feature set (rank, not magnitude).
concordance_tbl <- function(long) {
  pairs <- utils::combn(levels(long$method), 2, simplify = FALSE)
  purrr::map_dfr(pairs, function(p) {
    a <- dplyr::filter(long, method == p[1]); b <- dplyr::filter(long, method == p[2])
    m <- dplyr::inner_join(a, b, by = "feature", suffix = c("_a", "_b"))
    both <- m$sig_a & m$sig_b
    tibble::tibble(
      `Method A`   = p[1], `Method B` = p[2],
      `Both sig`   = sum(both, na.rm = TRUE),
      `Sign agree` = if (any(both)) sprintf("%.0f%%",
                       100 * mean(sign(m$effect_a[both]) == sign(m$effect_b[both]))) else "",
      Spearman     = sprintf("%.2f", suppressWarnings(
                       stats::cor(m$effect_a, m$effect_b,
                                  method = "spearman", use = "complete.obs")))
    )
  })
}

# Consensus table: union of features called by any method, one row per feature,
# each method shown as "q Dir", ordered by how many methods called it then q.
consensus_tbl <- function(long, top = 30) {
  sig_feats <- unique(dplyr::pull(dplyr::filter(long, sig), feature))
  if (length(sig_feats) == 0L) return(NULL)
  cells <- long |>
    dplyr::filter(feature %in% sig_feats) |>
    dplyr::mutate(cell = ifelse(is.na(qvalue), "",
                    paste0(fmt_q(qvalue), ifelse(Dir == "", "", paste0(" ", Dir)))))
  nsig <- cells |>
    dplyr::group_by(feature) |>
    dplyr::summarise(`n methods` = sum(sig, na.rm = TRUE),
                     min_q = suppressWarnings(min(qvalue, na.rm = TRUE)),
                     .groups = "drop")
  cells |>
    dplyr::select(feature, method, cell) |>
    tidyr::pivot_wider(names_from = method, values_from = cell) |>
    dplyr::left_join(nsig, by = "feature") |>
    dplyr::arrange(dplyr::desc(`n methods`), min_q) |>
    dplyr::mutate(Feature = feature) |>
    dplyr::select(Feature, dplyr::all_of(method_levels), `n methods`) |>
    utils::head(top)
}

sig_count <- function(long, m) sum(long$sig[long$method == m], na.rm = TRUE)

# Short, file-safe column tags per method for the wide export.
method_tags <- c("LinDA" = "LinDA", "MaAsLin3" = "MaAsLin3",
                 "ANCOM-BC2" = "ANCOM_BC2")

# Wide comparison table: one row per feature with full (untruncated) names and
# every method's effect, p, q and direction side by side, sorted so the
# multi-method consensus hits sit on top. This is the export-friendly view; the
# in-report consensus table is a trimmed version of the same information.
wide_results <- function(long) {
  w <- long |>
    dplyr::mutate(method = unname(method_tags[as.character(method)])) |>
    dplyr::select(feature, method, effect, pvalue, qvalue, Dir, sig)
  nsig <- w |>
    dplyr::group_by(feature) |>
    dplyr::summarise(`n methods` = sum(sig, na.rm = TRUE),
                     min_q = suppressWarnings(min(qvalue, na.rm = TRUE)),
                     .groups = "drop")
  wide <- w |>
    dplyr::select(-sig) |>
    tidyr::pivot_wider(names_from = method,
                       values_from = c(effect, pvalue, qvalue, Dir),
                       names_glue = "{method}_{.value}")
  ord <- unlist(lapply(unname(method_tags),
                       \(m) paste0(m, c("_effect", "_pvalue", "_qvalue", "_Dir"))))
  ord <- ord[ord %in% names(wide)]
  wide |>
    dplyr::left_join(nsig, by = "feature") |>
    dplyr::arrange(dplyr::desc(`n methods`), min_q) |>
    dplyr::select(Feature = feature, `n methods`, dplyr::all_of(ord))
}

# Write the wide tables as CSVs (always) and, if a writer package is available,
# as a single multi-sheet Excel workbook. openxlsx is preferred (no Java, adds
# a header filter, a frozen header row and a wide name column); writexl is a
# lighter fallback. Returns a one-line description of what was written.
write_da_exports <- function(sheets, out_dir) {
  for (nm in names(sheets))
    utils::write.csv(sheets[[nm]],
      file.path(out_dir, paste0("DA_comparison_", tolower(nm), "_wide.csv")),
      row.names = FALSE)
  xlsx_path <- file.path(out_dir, "DA_method_comparison.xlsx")
  if (requireNamespace("openxlsx", quietly = TRUE)) {
    wb <- openxlsx::createWorkbook()
    for (nm in names(sheets)) {
      openxlsx::addWorksheet(wb, nm)
      openxlsx::writeData(wb, nm, sheets[[nm]], withFilter = TRUE)
      openxlsx::freezePane(wb, nm, firstActiveRow = 2, firstActiveCol = 2)
      openxlsx::setColWidths(wb, nm, cols = 1, widths = 70)
    }
    openxlsx::saveWorkbook(wb, xlsx_path, overwrite = TRUE)
    sprintf("CSV files and an Excel workbook (%s)", basename(xlsx_path))
  } else if (requireNamespace("writexl", quietly = TRUE)) {
    writexl::write_xlsx(sheets, xlsx_path)
    sprintf("CSV files and an Excel workbook (%s)", basename(xlsx_path))
  } else {
    "CSV files only (install openxlsx or writexl for the Excel workbook)"
  }
}

# Per-method volcano panels, faceted (free x because effect scales differ).
vol_facet <- function(long) {
  v <- long |>
    dplyr::filter(!is.na(qvalue)) |>
    dplyr::mutate(Support = factor(ifelse(sig, "called", "ns"),
                                   levels = c("called", "ns")))
  lab <- v |>
    dplyr::filter(sig) |>
    dplyr::group_by(method) |>
    dplyr::arrange(qvalue, .by_group = TRUE) |>
    dplyr::slice_head(n = 5) |>
    dplyr::ungroup() |>
    dplyr::mutate(Label = stringr::str_trunc(feature, 30))
  ggplot(v, aes(effect, -log10(qvalue), colour = Support)) +
    geom_point(size = 1.3, alpha = 0.85) +
    ggrepel::geom_text_repel(data = lab, aes(label = Label), colour = "black",
      size = 2.1, box.padding = 0.5, max.overlaps = Inf,
      min.segment.length = 0, seed = 1) +
    geom_hline(yintercept = -log10(q_cut), linetype = 2, colour = "grey50") +
    geom_vline(xintercept = 0, linetype = 3, colour = "grey50") +
    facet_wrap(~ method, scales = "free_x", nrow = 1) +
    scale_colour_manual(values = c(called = "#C0392B", ns = "grey75")) +
    labs(x = "Coefficient (HRSSL vs NN), method-native scale",
         y = expression(-log[10]~q[BH]), colour = NULL)
}
```

# Differential abundance: pathways

```{r da-path, results='hide'}
path_long <- add_flags(dplyr::bind_rows(
  run_linda(path_cpm_f, meta),
  run_maaslin3(path_cpm_f, meta, "maaslin3_pathways", path_name_map),
  run_ancombc2(path_rpk_f, meta, path_name_map)
))

write.csv(path_long, file.path(out_dir, "DA_method_comparison_pathways.csv"),
          row.names = FALSE)

p_linda   <- sig_count(path_long, "LinDA")
p_m3      <- sig_count(path_long, "MaAsLin3")
p_ancombc <- sig_count(path_long, "ANCOM-BC2")
p_feat_nsig <- path_long |> dplyr::filter(sig) |> dplyr::count(feature)
p_all3  <- sum(p_feat_nsig$n == length(method_levels))
p_any   <- nrow(p_feat_nsig)
p_only1 <- sum(p_feat_nsig$n == 1)
```

At \(q < `r q_cut`\) the three methods call `r p_linda` (LinDA),
`r p_m3` (MaAsLin 3), and `r p_ancombc` (ANCOM-BC2) pathways. Across methods,
`r p_any` distinct pathways are called by at least one method, `r p_all3` by all
three, and `r p_only1` by only one.

```{r path-concordance}
concordance_tbl(path_long) |>
  kbl(booktabs = TRUE,
      caption = "Pairwise concordance, pathway layer. Both sig: features called by both methods at the threshold. Sign agree: proportion of those with matching direction. Spearman: rank correlation of effect over all shared features.") |>
  kable_styling(latex_options = c("hold_position"))
```

```{r path-consensus}
ct <- consensus_tbl(path_long)
if (is.null(ct)) {
  cat("No pathway reached the q threshold under any method.")
} else {
  ct |>
    kbl(booktabs = TRUE, longtable = TRUE,
        caption = "Pathways called by at least one method, ordered by number of methods then q-value. Each method cell is the q-value and direction (Up / Dn in HRSSL). Features agreed by all three are the defensible core. Full names for every feature are in the exported tables.") |>
    kable_styling(latex_options = c("repeat_header", "hold_position"), font_size = 8) |>
    column_spec(1, width = "11cm") |>
    kableExtra::landscape()
}
```

```{r path-volcano, fig.height=3.6, fig.cap="Per-method volcano, pathway layer. Called features (q < 0.25) in red; dashed line marks the FDR threshold; the five most significant per method are labelled. The x axis is free per panel because effect scales differ across methods."}
vol_facet(path_long)
```

# Differential abundance: KEGG orthologs

```{r da-ko, results='hide'}
ko_long <- add_flags(dplyr::bind_rows(
  run_linda(ko_cpm_f, meta),
  run_maaslin3(ko_cpm_f, meta, "maaslin3_kos", ko_name_map),
  run_ancombc2(ko_rpk_f, meta, ko_name_map)
))

write.csv(ko_long, file.path(out_dir, "DA_method_comparison_kos.csv"),
          row.names = FALSE)

k_linda   <- sig_count(ko_long, "LinDA")
k_m3      <- sig_count(ko_long, "MaAsLin3")
k_ancombc <- sig_count(ko_long, "ANCOM-BC2")
k_feat_nsig <- ko_long |> dplyr::filter(sig) |> dplyr::count(feature)
k_all3  <- sum(k_feat_nsig$n == length(method_levels))
k_any   <- nrow(k_feat_nsig)
k_only1 <- sum(k_feat_nsig$n == 1)
```

At \(q < `r q_cut`\) the three methods call `r k_linda` (LinDA),
`r k_m3` (MaAsLin 3), and `r k_ancombc` (ANCOM-BC2) KOs. Across methods,
`r k_any` distinct KOs are called by at least one method, `r k_all3` by all
three, and `r k_only1` by only one.

```{r ko-concordance}
concordance_tbl(ko_long) |>
  kbl(booktabs = TRUE,
      caption = "Pairwise concordance, KO layer. Columns as in the pathway table.") |>
  kable_styling(latex_options = c("hold_position"))
```

```{r ko-consensus}
ct <- consensus_tbl(ko_long)
if (is.null(ct)) {
  cat("No KO reached the q threshold under any method.")
} else {
  ct |>
    kbl(booktabs = TRUE, longtable = TRUE,
        caption = "KOs called by at least one method, ordered by number of methods then q-value. Columns as in the pathway consensus table. Full names for every feature are in the exported tables.") |>
    kable_styling(latex_options = c("repeat_header", "hold_position"), font_size = 8) |>
    column_spec(1, width = "11cm") |>
    kableExtra::landscape()
}
```

```{r ko-volcano, fig.height=3.6, fig.cap="Per-method volcano, KO layer. Conventions as in the pathway volcano."}
vol_facet(ko_long)
```

# Exported tables

The complete results, with full untruncated feature names and every method's
effect, p-value, q-value and direction side by side, are written to the output
directory for comparison outside this report. There is one row per feature,
sorted so the features called by the most methods sit on top; nothing is
thresholded out, so the full tested feature set is present.

```{r exports}
exported <- write_da_exports(
  list(Pathways = wide_results(path_long), KOs = wide_results(ko_long)),
  out_dir
)
```

This run wrote `r exported`. The wide tables are
`DA_comparison_pathways_wide.csv` and `DA_comparison_kos_wide.csv`; the
per-method long tables `DA_method_comparison_pathways.csv` and
`DA_method_comparison_kos.csv` are written as before.

# Interpretation

The pairwise concordance tables are the part to read first. All three methods
recover the reference frame that plain relative abundance distorts, each by its
own route, so the question is whether they still agree once each has made its
correction. High both-called counts with strong sign agreement and Spearman
correlation across the LinDA, MaAsLin 3 and ANCOM-BC2 pairs indicate that the
choice of correction is not driving the calls, and that the shared hits reflect
the underlying biology rather than any single method's reference assumption.

For the manuscript, the consensus tables give a graded hit list. Features called
by all three methods are the core claims, robust to the choice of compositional
correction. Features called by two carry moderate support. Features called by
only one method are the most likely to reflect a method-specific reference
effect rather than a stable biological difference, and are best reported as
exploratory or held back. With n = 12 per group, this cross-method agreement is
a more honest robustness statement than any single method's q-value, and it
pairs naturally with the axis-level directionality already in the main analysis.

# Session info

```{r sessioninfo, echo=FALSE}
sessionInfo()
```
