---
title: "HUMAnN3 Functional Analysis: HRSSL vs NN"
author: "Brad"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    keep_tex: false
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
geometry: margin=2.2cm
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, warning = FALSE, message = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5,
  fig.pos = "H", out.extra = "",
  dev = "cairo_pdf"
)
options(knitr.kable.NA = "")
```

```{r libs}
# Bioconductor / S4Vectors-touching packages FIRST so tidyverse can mask cleanly
suppressPackageStartupMessages({
  library(Maaslin2)
  library(vegan)
  library(compositions)   # clr() / ilr() if needed
  library(ggplot2)
  library(ggrepel)
  library(ggpubr)
  library(scales)
  library(kableExtra)
  library(patchwork)
  library(RColorBrewer)
  # tidyverse / dplyr LAST
  library(tidyverse)
})

# Explicit dplyr namespacing convention - use dplyr:: in code chunks
theme_set(theme_bw(base_size = 10) +
          theme(panel.grid.minor = element_blank(),
                strip.background = element_rect(fill = "grey95", colour = NA)))

group_colours <- c("HRSSL" = "#C0392B", "NN" = "#2E86AB")
```

```{r paths}
# ---- EDIT THESE PATHS --------------------------------------------------------
proc_dir <- "humann3_processed"

# HUMAnN3 already names pathway features ("PWY-ID: name"); the KO tables are
# bare K-IDs out of the regrouping step and are passed through
# humann_rename_table to attach KEGG ortholog names. Both the CPM and RPK KO
# inputs below are the _named_ outputs of that step, so KO IDs carry names
# consistently across the two layers.
path_cpm_unstrat_file <- file.path(proc_dir, "path_split_cpm",
                                   "all_pathabundance_cpm_named_unstratified.tsv")
path_cpm_strat_file   <- file.path(proc_dir, "path_split_cpm",
                                   "all_pathabundance_cpm_stratified.tsv")
ko_cpm_unstrat_file   <- file.path(proc_dir, "ko_split_cpm",
                                   "all_ko_cpm_named_unstratified.tsv")
path_rpk_unstrat_file <- file.path(proc_dir, "path_split_rpk",
                                   "all_pathabundance_unstratified.tsv")
ko_rpk_unstrat_file   <- file.path(proc_dir, "ko_split_rpk",
                                   "all_ko_unstratified_named.tsv")

metadata_file         <- "nnandhrsslmetadata.csv"   # SampleID, Group  (HRSSL / NN)

out_dir <- "humann3_analysis_out"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)
# -----------------------------------------------------------------------------
```

# Overview

This report processes HUMAnN3 functional profiles from 24 stool metagenomes
(12 high-risk sessile serrated lesion patients, HRSSL; 12 no-neoplasia
controls, NN) and addresses two questions:

\begin{enumerate}
  \item Which MetaCyc pathways and KEGG orthologs (KOs) are differentially
        abundant between groups?
  \item Which of those features map onto the functional axes implicated in
        sessile serrated neoplasia, dysbiosis, and colorectal carcinogenesis?
\end{enumerate}

The analytical strategy uses MaAsLin2 as the DA framework: a linear model on
log-transformed CPM abundance with `Group` and per-layer feature richness as
fixed effects, and Benjamini-Hochberg FDR control. CPM is already a
relative-abundance scaling, so MaAsLin2's own normalisation is disabled.
Richness is centred and scaled and entered as a covariate to absorb the
closure-driven component that arises from the lower feature detection in HRSSL
samples; the `Group` coefficient is therefore adjusted for richness before
FDR correction. Significance is taken at MaAsLin2's default FDR threshold of
\(q < 0.25\) to preserve sensitivity in this small cohort (n = 12/group).

# Data ingestion

```{r read-helper}
# HUMAnN tables have one feature column then sample columns. Read robustly,
# strip "# " comment from header, and coerce to matrix with rownames = features.
read_humann <- function(path) {
  tab <- read.delim(path, sep = "\t", header = TRUE,
                    check.names = FALSE, comment.char = "")
  names(tab)[1] <- "Feature"
  # Defensive: an earlier (broken) rename pass may have appended ": NO_NAME"
  # to entries that were already named. Strip it so feature IDs match across
  # cached files and the live ID/name format.
  tab$Feature <- sub(":\\s*NO_NAME\\s*$", "", tab$Feature)
  feat_col <- tab$Feature
  mat <- as.matrix(tab[, -1, drop = FALSE])
  rownames(mat) <- feat_col
  mat
}

drop_unmapped <- function(mat) {
  # Pathway tables carry UNMAPPED + UNINTEGRATED; KO / gene-family tables carry
  # UNMAPPED + UNGROUPED. Strip both conventions (and their |unclassified
  # stratified variants) so neither catch-all bin is treated as a real feature.
  keep <- !rownames(mat) %in% c("UNMAPPED", "UNINTEGRATED", "UNGROUPED",
                                "UNMAPPED|unclassified",
                                "UNINTEGRATED|unclassified",
                                "UNGROUPED|unclassified")
  mat[keep, , drop = FALSE]
}
```

```{r read-data}
path_cpm <- read_humann(path_cpm_unstrat_file)
ko_cpm   <- read_humann(ko_cpm_unstrat_file)
path_rpk <- read_humann(path_rpk_unstrat_file)
ko_rpk   <- read_humann(ko_rpk_unstrat_file)
path_strat <- read_humann(path_cpm_strat_file)

# Keep UNMAPPED/UNINTEGRATED for QC, exclude them from the analytical matrices
qc_path    <- path_cpm
path_cpm   <- drop_unmapped(path_cpm)
ko_cpm     <- drop_unmapped(ko_cpm)
path_rpk   <- drop_unmapped(path_rpk)
ko_rpk     <- drop_unmapped(ko_rpk)
# Stratified file also contains UNMAPPED|... and UNINTEGRATED|... rows; drop them.
path_strat <- path_strat[!grepl("^(UNMAPPED|UNINTEGRATED)", rownames(path_strat)), ,
                         drop = FALSE]

# --- Pathway-naming diagnostic ----------------------------------------------
# Modern HUMAnN3 emits pathway IDs of the form "PWY-ID: name". Check that
# the loaded matrix has names attached (otherwise the targeted axes still
# work via MetaCyc ID matching, but the top-25 table will be less readable).
named_frac <- mean(grepl(": ", rownames(path_cpm)))
if (named_frac < 0.5) {
  message(sprintf(
    "Only %.0f%% of pathway IDs carry human-readable names. ",
    100 * named_frac),
    "Targeted axes match MetaCyc IDs directly, so analysis is unaffected; ",
    "the top-25 table will show bare IDs.")
}

meta <- read.csv(metadata_file, stringsAsFactors = FALSE)
stopifnot(all(c("SampleID", "Group") %in% names(meta)))

# Clean Group values: strip whitespace, normalise case, then factor.
# Any value not in {NN, HRSSL} becomes NA at the factor() step; we then
# explicitly drop NA rows and warn about what was lost.
meta$SampleID <- trimws(as.character(meta$SampleID))
meta$Group    <- toupper(trimws(as.character(meta$Group)))

unexpected_grp <- setdiff(unique(meta$Group), c("NN", "HRSSL", NA, ""))
if (length(unexpected_grp)) {
  warning("Group values not in {NN, HRSSL} will be dropped: ",
          paste(unexpected_grp, collapse = ", "))
}

meta$Group <- factor(meta$Group, levels = c("NN", "HRSSL"))

# Drop rows with missing SampleID or Group
n_before <- nrow(meta)
meta <- meta[!is.na(meta$Group) & nzchar(meta$SampleID), , drop = FALSE]
if (nrow(meta) < n_before) {
  message(sprintf("Dropped %d metadata rows with missing SampleID or Group.",
                  n_before - nrow(meta)))
}

rownames(meta) <- meta$SampleID

# Restrict to samples present in BOTH metadata and every matrix layer.
# (Previously this chunk stopped on any mismatch; now it intersects and warns.)
all_mats   <- list(path_cpm = path_cpm, ko_cpm = ko_cpm,
                   path_rpk = path_rpk, ko_rpk = ko_rpk,
                   qc_path  = qc_path,  path_strat = path_strat)
common <- Reduce(intersect, c(list(meta$SampleID),
                              lapply(all_mats, colnames)))

dropped_meta <- setdiff(meta$SampleID, common)
dropped_mat  <- setdiff(Reduce(union, lapply(all_mats, colnames)), common)
if (length(dropped_meta)) {
  warning("Samples in metadata but absent from one or more matrices: ",
          paste(dropped_meta, collapse = ", "))
}
if (length(dropped_mat)) {
  warning("Samples in matrices but absent from metadata: ",
          paste(head(dropped_mat, 10), collapse = ", "),
          if (length(dropped_mat) > 10) sprintf(" (+%d more)",
                                                length(dropped_mat) - 10))
}
stopifnot(length(common) >= 4)  # need at least a few samples per group

samples    <- common
meta       <- meta[samples, , drop = FALSE]
path_cpm   <- path_cpm[,   samples, drop = FALSE]
ko_cpm     <- ko_cpm[,     samples, drop = FALSE]
path_rpk   <- path_rpk[,   samples, drop = FALSE]
ko_rpk     <- ko_rpk[,     samples, drop = FALSE]
qc_path    <- qc_path[,    samples, drop = FALSE]
path_strat <- path_strat[, samples, drop = FALSE]
```

```{r data-summary}
data.frame(
  Layer    = c("MetaCyc pathways (CPM)", "MetaCyc pathways (RPK)",
               "KEGG orthologs (CPM)",   "KEGG orthologs (RPK)",
               "Stratified pathways (CPM)"),
  Features = c(nrow(path_cpm), nrow(path_rpk),
               nrow(ko_cpm),   nrow(ko_rpk),
               nrow(path_strat)),
  Samples  = c(ncol(path_cpm), ncol(path_rpk),
               ncol(ko_cpm),   ncol(ko_rpk),
               ncol(path_strat))
) |>
  kbl(booktabs = TRUE, caption = "Input data dimensions after join, normalise, name, split.") |>
  kable_styling(latex_options = c("hold_position"))
```

# Sample QC

```{r qc-mapping}
unmap_row     <- qc_path["UNMAPPED", , drop = TRUE]
unint_row     <- qc_path["UNINTEGRATED", , drop = TRUE]
mapped_pwy    <- 1e6 - unmap_row - unint_row
detected_pwy  <- colSums(path_cpm > 0)
detected_ko   <- colSums(ko_cpm > 0)

qc <- data.frame(
  SampleID     = samples,
  Group        = meta$Group,
  Unmapped_CPM = unmap_row,
  Unintegrated_CPM = unint_row,
  Mapped_pwy_CPM = mapped_pwy,
  Pathways_detected = detected_pwy,
  KOs_detected      = detected_ko
)
```

```{r qc-tables}
qc |>
  dplyr::group_by(Group) |>
  dplyr::summarise(
    n          = dplyr::n(),
    pwy_median = round(median(Pathways_detected)),
    pwy_range  = sprintf("%d - %d", min(Pathways_detected), max(Pathways_detected)),
    ko_median  = round(median(KOs_detected)),
    ko_range   = sprintf("%d - %d", min(KOs_detected), max(KOs_detected)),
    unmap_med  = round(median(Unmapped_CPM)),
    .groups    = "drop"
  ) |>
  kbl(booktabs = TRUE,
      col.names = c("Group", "n", "Pathways (median)", "Pathways (range)",
                    "KOs (median)", "KOs (range)", "Unmapped CPM (median)"),
      caption = "Per-group feature detection summary.") |>
  kable_styling(latex_options = c("hold_position", "scale_down"))
```

```{r qc-plots, fig.height=3.5, fig.cap="Pathway and KO detection counts by group; high UNMAPPED CPM in any sample flags poor reference coverage."}
p1 <- ggplot(qc, aes(Group, Pathways_detected, fill = Group)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, size = 1.6, alpha = 0.8) +
  scale_fill_manual(values = group_colours, guide = "none") +
  labs(y = "Pathways detected", x = NULL)

p2 <- ggplot(qc, aes(Group, KOs_detected, fill = Group)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, size = 1.6, alpha = 0.8) +
  scale_fill_manual(values = group_colours, guide = "none") +
  labs(y = "KOs detected", x = NULL)

p3 <- ggplot(qc, aes(Group, Unmapped_CPM / 1e4, fill = Group)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.8) +
  geom_jitter(width = 0.12, size = 1.6, alpha = 0.8) +
  scale_fill_manual(values = group_colours, guide = "none") +
  labs(y = "UNMAPPED (% of CPM)", x = NULL)

p1 + p2 + p3
```

# Ordination & community-level differences

Two complementary views of between-sample structure: Bray-Curtis dissimilarity
on CPM-scaled pathway abundance (relative-abundance space) and Aitchison
distance on CLR-transformed RPK (compositional space). PERMANOVA on the same
distance matrices tests whether `Group` explains a non-trivial fraction of
the between-sample variance.

```{r ordination, fig.height=4, fig.cap="Bray-Curtis NMDS on pathway CPM (left) and PCA on CLR-transformed pathway RPK (right). PERMANOVA reported below."}
# Bray-Curtis on TSS-scaled pathway abundance (HUMAnN CPM == TSS x 1e6)
bray <- vegan::vegdist(t(path_cpm), method = "bray")
set.seed(1)
nmds <- vegan::metaMDS(bray, k = 2, trymax = 100, trace = 0)
nmds_df <- data.frame(NMDS1 = nmds$points[,1], NMDS2 = nmds$points[,2],
                      Group = meta$Group, SampleID = samples)

# Aitchison (CLR-based) using RPK with pseudo-count
pc <- 0.5
path_clr <- t(apply(path_rpk + pc, 2, function(x) log(x) - mean(log(x))))
# rows = samples, cols = features
pca <- prcomp(path_clr, center = TRUE, scale. = FALSE)
pve <- (pca$sdev^2) / sum(pca$sdev^2)
pca_df <- data.frame(PC1 = pca$x[,1], PC2 = pca$x[,2],
                     Group = meta$Group, SampleID = samples)

# PERMANOVA
set.seed(1)
perm_bray <- vegan::adonis2(t(path_cpm) ~ Group, data = meta,
                            method = "bray", permutations = 999)
set.seed(1)
ait_dist <- dist(path_clr, method = "euclidean")
perm_ait <- vegan::adonis2(ait_dist ~ Group, data = meta, permutations = 999)

g_nmds <- ggplot(nmds_df, aes(NMDS1, NMDS2, colour = Group)) +
  geom_point(size = 2.4) +
  stat_ellipse(level = 0.8, linewidth = 0.6) +
  scale_colour_manual(values = group_colours) +
  labs(subtitle = sprintf("Bray-Curtis NMDS, stress = %.3f", nmds$stress))

g_pca <- ggplot(pca_df, aes(PC1, PC2, colour = Group)) +
  geom_point(size = 2.4) +
  stat_ellipse(level = 0.8, linewidth = 0.6) +
  scale_colour_manual(values = group_colours) +
  labs(x = sprintf("PC1 (%.1f%%)", 100 * pve[1]),
       y = sprintf("PC2 (%.1f%%)", 100 * pve[2]),
       subtitle = "CLR-PCA (Aitchison)")

(g_nmds | g_pca) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

```{r permanova-table}
permtab <- data.frame(
  Distance = c("Bray-Curtis (CPM)", "Aitchison (CLR-RPK)"),
  R2       = c(perm_bray$R2[1], perm_ait$R2[1]),
  F        = c(perm_bray$F[1],  perm_ait$F[1]),
  p        = c(perm_bray$`Pr(>F)`[1], perm_ait$`Pr(>F)`[1])
)
permtab |>
  dplyr::mutate(R2 = sprintf("%.3f", R2),
                F  = sprintf("%.2f", F),
                p  = sprintf("%.3f", p)) |>
  kbl(booktabs = TRUE, caption = "PERMANOVA on group (999 permutations).") |>
  kable_styling(latex_options = c("hold_position"))
```

\clearpage

# Prevalence filtering

Features detected (>0 CPM) in at least 25% of samples *in either group* are
kept. This removes singletons that destabilise the linear model but preserves
features specific to one group.

```{r prevalence}
# Keep features detected (>0) in >=25% of samples in at least one group.
prev_filter <- function(mat, groups, min_prev = 0.25) {
  grp <- split(colnames(mat), groups)
  keep <- vapply(rownames(mat), function(f) {
    any(vapply(grp, function(s) mean(mat[f, s] > 0) >= min_prev, logical(1)))
  }, logical(1))
  mat[keep, , drop = FALSE]
}

path_cpm_f <- prev_filter(path_cpm, meta$Group)
path_rpk_f <- prev_filter(path_rpk, meta$Group)
ko_cpm_f   <- prev_filter(ko_cpm,   meta$Group)
ko_rpk_f   <- prev_filter(ko_rpk,   meta$Group)

# Per-sample feature richness on the filtered layers. Pathway richness is
# used as the covariate in the pathway MaAsLin2 run; KO richness is used in
# the KO run. Both are centred and scaled before being added to the metadata
# so the richness coefficient is on a standardised scale and comparable
# across layers. This absorbs the closure-driven component that arises from
# the lower feature detection in HRSSL samples (see section 6.2).
sample_richness <- colSums(path_cpm_f > 0)   # pathway richness (kept for QC text)
pwy_rich <- scale(sample_richness)[, 1]       # centred/scaled pathway richness
ko_rich  <- scale(colSums(ko_cpm_f > 0))[, 1] # centred/scaled KO richness

# Attach both richness vectors to a shared metadata frame so MaAsLin2 can
# access them. (We pass layer-specific metadata to each run below.)
meta_pwy <- meta
meta_pwy$richness <- pwy_rich[rownames(meta_pwy)]

meta_ko <- meta
meta_ko$richness <- ko_rich[rownames(meta_ko)]

data.frame(
  Layer = c("Pathways", "KOs"),
  Before = c(nrow(path_cpm), nrow(ko_cpm)),
  After  = c(nrow(path_cpm_f), nrow(ko_cpm_f))
) |>
  kbl(booktabs = TRUE, caption = "Features before/after >=25% prevalence in either group.") |>
  kable_styling(latex_options = c("hold_position"))
```

# Differential abundance: pathways

## MaAsLin2

Linear model on log-transformed CPM pathway abundance, with `Group` and
per-sample pathway richness (centred and scaled) as fixed effects (reference
NN) and Benjamini-Hochberg FDR control. CPM is already a per-sample
relative-abundance scaling, so MaAsLin2's internal normalisation is set to
`NONE`. The richness covariate absorbs the closure-driven component arising
from lower feature detection in HRSSL samples, so the `Group` coefficient
reflects between-group differences net of richness. Significance is taken at
MaAsLin2's default FDR threshold of *q* < 0.25 to preserve sensitivity in
this small cohort (n = 12/group). Output is suppressed; results are
summarised in section 6.2.

```{r maaslin2-path, results='hide'}
maas_path_dir <- file.path(out_dir, "maaslin2_pathways")
unlink(maas_path_dir, recursive = TRUE)

# Model choice follows the bioBakery guidance (Mallick, MaAsLin2 forum): the
# model, normalisation, and transform must be matched to the input type.
#   - HUMAnN CPM (and RPK) are continuous values, not integer counts, so the
#     count models (NEGBIN, ZINB) and the count-only normalisations (TMM, CSS)
#     do not apply. The valid non-count models are LM and CPLM.
#   - CPM is already a per-sample relative-abundance scaling, so normalisation
#     is NONE; TSS here would double-normalise.
#   - LM is the only model that accepts both positive and negative values, so
#     it is valid with transform = LOG (CPLM would instead require
#     transform = NONE on strictly positive data, with a log-link coefficient).
#     LM keeps the Group coefficient interpretable on the log-abundance scale.
# Net: LM + normalization NONE + transform LOG on CPM.
# Richness (centred/scaled pathway richness) is added as a second fixed effect
# to absorb the closure-driven inflation caused by lower feature detection in
# HRSSL. The Group coefficient is therefore adjusted for richness.
maas_input  <- as.data.frame(t(path_cpm_f))             # samples = rows, CPM
maas_meta   <- meta_pwy[rownames(maas_input), , drop = FALSE]

maas_path <- Maaslin2::Maaslin2(
  input_data       = maas_input,
  input_metadata   = maas_meta,
  output           = maas_path_dir,
  fixed_effects    = c("Group", "richness"),
  reference        = c("Group,NN"),
  normalization    = "NONE",         # CPM already relative-abundance scaled
  transform        = "LOG",
  analysis_method  = "LM",
  correction       = "BH",
  min_abundance    = 0,
  min_prevalence   = 0,        # we filtered already
  max_significance = 0.25,
  plot_heatmap     = FALSE,
  plot_scatter     = FALSE,
  standardize      = FALSE
)

maas_path_res <- maas_path$results |>
  dplyr::filter(metadata == "Group") |>
  dplyr::transmute(
    Feature = feature,
    Coef    = coef,
    StdErr  = stderr,
    p_value = pval,
    q_value = qval,
    N_nonzero = N.not.zero
  ) |>
  dplyr::arrange(q_value)

# MaAsLin2 sanitises feature names via make.names(). Map them back to the
# original pathway IDs so the displayed names, the targeted-axis matching, and
# the stratified-layer join all use the real "PWY-ID: name" strings that match
# the abundance-matrix rownames.
path_name_map <- stats::setNames(rownames(path_cpm_f),
                                 make.names(rownames(path_cpm_f)))
maas_path_res$Feature <- dplyr::coalesce(
  unname(path_name_map[maas_path_res$Feature]), maas_path_res$Feature)
```

## Top hits

```{r overlap-path}
combo_path <- maas_path_res |>
  dplyr::mutate(
    sig_maaslin = !is.na(q_value) & q_value < 0.25,
    Direction = dplyr::case_when(
      !is.na(Coef) & Coef > 0 ~ "Up in HRSSL",
      !is.na(Coef) & Coef < 0 ~ "Down in HRSSL",
      TRUE ~ NA_character_
    )
  )

write.csv(combo_path, file.path(out_dir, "DA_pathways_maaslin.csv"), row.names = FALSE)

n_maas <- sum(combo_path$sig_maaslin, na.rm = TRUE)

# Quantify the richness asymmetry behind the closure caveat, for the text.
rich_lm   <- summary(stats::lm(sample_richness ~ meta$Group))
rich_nn   <- round(median(sample_richness[meta$Group == "NN"]))
rich_hr   <- round(median(sample_richness[meta$Group == "HRSSL"]))
rich_p    <- rich_lm$coefficients["meta$GroupHRSSL", "Pr(>|t|)"]
rich_p_fmt <- ifelse(rich_p < 1e-3, formatC(rich_p, format = "e", digits = 1),
                     sprintf("%.3f", rich_p))
```

MaAsLin2 (q < 0.25) yielded `r n_maas` differentially abundant pathways.

HRSSL samples carry substantially lower feature richness than NN (median
`r rich_hr` vs `r rich_nn` detected pathways; *p* = `r rich_p_fmt`). Because
CPM is a relative-abundance scaling, this asymmetry can redistribute proportion
onto surviving features and inflate apparent between-group differences — the
compositional closure problem. The richness covariate entered into the model
(centred and scaled pathway richness) absorbs this component directly, so the
`Group` coefficient reflects differences net of richness. Features in this list
are therefore less exposed to closure artefact than they would be in an
unadjusted model; the axis-level directionality in section 8 provides a
complementary view of coherent shifts across related features.

```{r top-path-table}
# Dir column gives the sign of the MaAsLin2 Group coefficient as Up / Dn /
# blank (Up = higher in HRSSL relative to NN).
dir_token <- function(x) dplyr::case_when(
  is.na(x)  ~ "",
  x > 0     ~ "Up",
  x < 0     ~ "Dn",
  TRUE      ~ ""
)

# Display q-values at 3 significant figures in fixed notation for q >= 1e-3
# (typical BH-adjusted range for this cohort) and scientific only for very
# small values. The previous `format="e", digits=1` rounding squashed any q
# in [0.195, 0.205] to "2.0e-01", hiding the real BH plateau structure that
# n=12/group produces on this dataset.
fmt_q <- function(q) {
  ifelse(is.na(q), "",
         ifelse(q >= 1e-3,
                sprintf("%.3f", q),
                formatC(q, format = "e", digits = 2)))
}

top_path <- combo_path |>
  dplyr::filter(sig_maaslin) |>
  dplyr::arrange(dplyr::coalesce(q_value, 1)) |>
  dplyr::transmute(
    Pathway = stringr::str_trunc(Feature, 55),
    Dir = dir_token(Coef),
    Coef = ifelse(is.na(Coef), "", sprintf("%+.2f", Coef)),
    q     = fmt_q(q_value),
    `N nonzero` = N_nonzero
  ) |>
  utils::head(25)

top_path |>
  kbl(booktabs = TRUE, longtable = TRUE,
      caption = "Top 25 differentially abundant pathways (MaAsLin2, q < 0.25), sorted by q-value. Dir gives the direction of the Group coefficient (Up / Dn in HRSSL relative to NN). Coef is the MaAsLin2 log-CPM coefficient; q is the BH-adjusted p-value; N nonzero is the number of samples in which the feature was detected.") |>
  kable_styling(latex_options = c("repeat_header", "hold_position"),
                font_size = 8) |>
  column_spec(1, width = "9cm")
```

```{r volcano-path, fig.height=4.2, fig.cap="MaAsLin2 volcano. Features called at q < 0.25 are highlighted; the dashed line marks the FDR threshold. The six most significant called features are labelled."}
vol_df <- combo_path |>
  dplyr::filter(!is.na(q_value)) |>
  dplyr::mutate(
    Support = ifelse(sig_maaslin, "q < 0.25", "ns"),
    Support = factor(Support, levels = c("q < 0.25", "ns"))
  )

# Label the most significant called features, cap at 6.
label_df <- vol_df |>
  dplyr::filter(sig_maaslin) |>
  dplyr::arrange(q_value) |>
  utils::head(6) |>
  dplyr::mutate(Label = stringr::str_trunc(Feature, 40))

ggplot(vol_df, aes(Coef, -log10(q_value), colour = Support)) +
  geom_point(size = 1.5, alpha = 0.85) +
  ggrepel::geom_text_repel(data = label_df,
                           aes(label = Label),
                           size = 2.6,
                           box.padding = 0.6,
                           point.padding = 0.3,
                           force = 4, force_pull = 0.5,
                           max.overlaps = Inf,
                           min.segment.length = 0,
                           seed = 1) +
  geom_hline(yintercept = -log10(0.25), linetype = 2, colour = "grey50") +
  geom_vline(xintercept = 0, linetype = 3, colour = "grey50") +
  scale_colour_manual(
    values = c("q < 0.25" = "#C0392B",   # red - called
               "ns"       = "grey75"),
    drop = FALSE) +
  labs(x = "MaAsLin2 coefficient (log CPM, HRSSL vs NN)",
       y = expression(-log[10]~q[BH]))
```

# Differential abundance: KEGG orthologs

```{r maaslin2-ko, results='hide'}
maas_ko_dir <- file.path(out_dir, "maaslin2_kos")
unlink(maas_ko_dir, recursive = TRUE)

# Same model family as the pathway run (see the model-choice note there):
# CPM is non-count, already normalised, so LM + normalization NONE + LOG.
# KO richness (centred/scaled) is the covariate here, matching the layer.
maas_ko_input <- as.data.frame(t(ko_cpm_f))
maas_ko_meta  <- meta_ko[rownames(maas_ko_input), , drop = FALSE]

maas_ko <- Maaslin2::Maaslin2(
  input_data       = maas_ko_input,
  input_metadata   = maas_ko_meta,
  output           = maas_ko_dir,
  fixed_effects    = c("Group", "richness"),
  reference        = c("Group,NN"),
  normalization    = "NONE",         # CPM already relative-abundance scaled
  transform        = "LOG",
  analysis_method  = "LM",
  correction       = "BH",
  min_abundance    = 0,
  min_prevalence   = 0,
  max_significance = 0.25,
  plot_heatmap     = FALSE,
  plot_scatter     = FALSE,
  standardize      = FALSE
)

maas_ko_res <- maas_ko$results |>
  dplyr::filter(metadata == "Group") |>
  dplyr::transmute(Feature = feature, Coef = coef,
                   p_value = pval, q_value = qval,
                   N_nonzero = N.not.zero) |>
  dplyr::arrange(q_value)

# Recover original KO IDs from MaAsLin2's make.names() sanitisation.
ko_name_map <- stats::setNames(rownames(ko_cpm_f),
                               make.names(rownames(ko_cpm_f)))
maas_ko_res$Feature <- dplyr::coalesce(
  unname(ko_name_map[maas_ko_res$Feature]), maas_ko_res$Feature)
```

```{r overlap-ko}
combo_ko <- maas_ko_res |>
  dplyr::mutate(
    sig_maaslin = !is.na(q_value) & q_value < 0.25,
    Direction = dplyr::case_when(
      !is.na(Coef) & Coef > 0 ~ "Up in HRSSL",
      !is.na(Coef) & Coef < 0 ~ "Down in HRSSL",
      TRUE ~ NA_character_
    )
  )

write.csv(combo_ko, file.path(out_dir, "DA_kos_maaslin.csv"), row.names = FALSE)
```

```{r top-ko-table}
combo_ko |>
  dplyr::filter(sig_maaslin) |>
  dplyr::arrange(dplyr::coalesce(q_value, 1)) |>
  dplyr::transmute(
    KO = stringr::str_trunc(Feature, 55),
    Dir = dir_token(Coef),
    Coef = ifelse(is.na(Coef), "", sprintf("%+.2f", Coef)),
    q     = fmt_q(q_value),
    `N nonzero` = N_nonzero
  ) |>
  utils::head(25) |>
  kbl(booktabs = TRUE, longtable = TRUE,
      caption = "Top 25 differentially abundant KOs (MaAsLin2, q < 0.25), sorted by q-value. Columns as in the pathway table.") |>
  kable_styling(latex_options = c("repeat_header", "hold_position"),
                font_size = 8) |>
  column_spec(1, width = "9cm")
```

# Targeted analysis: SSL / dysbiosis / CRC functional axes

This section interrogates a curated set of MetaCyc / KO patterns implicated
in colorectal carcinogenesis, with emphasis on the serrated neoplasia
pathway. We report directionality (HRSSL vs NN) for every axis member
regardless of FDR, because under-powered targeted contrasts are still
informative as biological priors.

```{r curated-axes}
# Each axis carries (a) a list of MetaCyc pathway IDs / KO IDs that anchor the
# axis biologically, and (b) a name-pattern regex as a backup for installs
# where humann_rename_table resolved human-readable names. Features are
# matched if EITHER pattern hits.
#
# MetaCyc pathway IDs follow either "<NAME>-PWY" or "PWY-<NNNN>" conventions;
# IDs are stable across HUMAnN releases.

pathway_axes <- tibble::tribble(
  ~Axis,                                 ~ID_pattern,                                                                                                ~Name_pattern,
  "Butyrate / SCFA production",          "^(CENTFERM-PWY|PWY-5022|PWY-5676|PWY-5677|P162-PWY|CODH-PWY|PWY-5482|PWY-5100|ACETATEUTIL-PWY|PWY-7790)(:|$)", "butanoate|butyrate|butyryl|acetyl-CoA ferment",
  "Sulfur / hydrogen sulfide",           "^(SO4ASSIM-PWY|PWY-5304|P222-PWY|PWY-5350|PWY-6683|DISSULFRED-PWY|PWY-5340|METHIONINE-DEG1-PWY|CYSDEGSE-PWY)(:|$)", "sulfate|sulfide|sulfur|sulfite|cysteine degrad|methionine degrad",
  "Polyamine biosynthesis",              "^(POLYAMSYN-PWY|POLYAMINSYN3-PWY|PWY-46|PWY-43|PWY-6305|PWY-6562|BSUBPOLYAMSYN-PWY|ARGSPECAT-PWY)(:|$)",     "putrescine|spermidine|spermine|polyamine|ornithine decarbox",
  "Folate / one-carbon",                 "^(1CMET2-PWY|PWY-3841|PWY-2201|PWY-6614|FOLSYN-PWY|PWY-2161|PWY-6147)(:|$)",                                "folate|tetrahydrofolate|one-carbon",
  "B12 / cobalamin",                     "^(COBALSYN-PWY|PWY-5509|PWY-5507|PWY-6269)(:|$)",                                                            "cobalamin|B12|adenosylcobalamin",
  "Other B-vitamins (B1/B2/B5/B7)",      "^(RIBOSYN2-PWY|PWY-6168|THISYN-PWY|THISYNARA-PWY|BIOTIN-BIOSYNTHESIS-PWY|PWY-5005|PANTO-PWY|COA-PWY|COA-PWY-1)(:|$)", "riboflavin|biotin|thiamine|pantothenate|coenzyme A",
  "LPS / KDO / ECA (Proteobacteria)",    "^(KDO-LIPASYN-PWY|LPSSYN-PWY|NAGLIPASYN-PWY|ECASYN-PWY|PWY-1269|PWY-6385)(:|$)",                              "lipopolysaccharide|KDO|lipid IVA|lipid A|enterobacterial common",
  "Mucin / glycan utilisation",          "^(FUCCAT-PWY|LACTOSECAT-PWY|PWY-3801|PWY-7242|GALACT-GLUCUROCAT-PWY|PWY-6507|PWY-6317)(:|$)",                  "fucose|sialic|N-acetylgalactos|N-acetylglucos|galactos.*degrad|glucuron",
  "Arginine / ornithine catabolism",     "^(ARGDEG-PWY|AST-PWY|ORNARGDEG-PWY|ORNDEG-PWY|ARG-PRO-PWY|ARGDEG-V-PWY|PWY-5004|PWY0-823)(:|$)",              "arginine degrad|ornithine degrad",
  "BCAA biosynthesis (Val/Leu/Ile)",     "^(VALSYN-PWY|ILEUSYN-PWY|LEUSYN-PWY|BRANCHED-CHAIN-AA-SYN-PWY|PWY-5101|PWY-5103)(:|$)",                       "valine biosynth|leucine biosynth|isoleucine biosynth|branched-chain",
  "Histidine biosynthesis",              "^(HISTSYN-PWY|HISDEG-PWY)(:|$)",                                                                              "histidine biosynth",
  "Nitrogen / denitrification",          "^(DENITRIFICATION-PWY|PWY-5675|PWY-381|PWY490-3|PWY-6964|AMMOXID-PWY|N2FIX-PWY)(:|$)",                         "nitrate|nitrite|nitric oxide|denitrif|nitrogen fix"
)

ko_axes <- tibble::tribble(
  ~Axis,                                   ~ID_pattern,                                                                       ~Name_pattern,
  "Butyrate kinase / butyryl-CoA / Bcd",   "^(K00929|K01034|K01035|K00248|K01615|K00074|K01034|K01615)(:|$)",                  "butyrate kinase|butyryl-CoA|butanoyl-CoA",
  "Bile acid 7alpha-dehydroxylation",      "^(K15868|K15869|K15870|K15871|K15872|K15873|K15874|K15875)(:|$)",                  "7alpha|7-alpha|bile acid",
  "Dissimilatory sulfite reductase (dsr)", "^(K11180|K11181|K00958|K00394|K00395)(:|$)",                                       "dissimilatory sulfite|sulfite reductase|sulfate adenylyltransferase",
  "Cysteine -> H2S (gamma-lyase etc.)",    "^(K01758|K17217|K14155|K01760|K01738)(:|$)",                                       "cystathionine|cysteine.*lyase|cysteine synthase",
  "Ornithine / polyamine biosynthesis",    "^(K01581|K01582|K01480|K00797|K00657)(:|$)",                                       "ornithine decarbox|spermidine synthase|spermine synthase|putrescine",
  "Colibactin (clb / pks island)",         "^(K23173|K23174|K23175|K23176|K23177|K23178|K23179|K23180|K23181|K23182|K23183|K23184|K23185|K23186|K23187|K23188|K23189|K23190)(:|$)", "colibactin|clbA|clbB|clbC|clbD|clbE|clbF|clbG|clbH|clbI|clbJ|clbK|clbL|clbM|clbN|clbO|clbP|clbQ|clbR|clbS",
  "Mucin sulfatase / sialidase",           "^(K12373|K01186|K01193|K01205|K12111)(:|$)",                                       "sulfatase|sialidase|fucosidase",
  "Folate biosynthesis (folA / metF)",     "^(K00287|K00297|K00548|K00549|K01098|K11754|K13939)(:|$)",                         "dihydrofolate|methylenetetrahydrofolate|folate",
  "Oxidative stress (catalase / SOD)",     "^(K03781|K04565|K00518|K00432|K03671|K07217)(:|$)",                                "catalase|superoxide dismutase|alkyl hydroperoxide|thioredoxin"
)

# Helper: match feature names against EITHER an ID pattern OR a name pattern
match_axis <- function(features, id_pat, name_pat) {
  feats <- features
  id_hits   <- grepl(id_pat,   feats, perl = TRUE)
  name_hits <- grepl(name_pat, feats, perl = TRUE, ignore.case = TRUE)
  feats[id_hits | name_hits]
}

axis_summary <- function(combo, mat_cpm, axes, layer) {
  feat_universe <- rownames(mat_cpm)
  # Match feature names case-sensitively, but join against `combo` via
  # make.names() so the filter is robust to MaAsLin2's name sanitisation
  # regardless of whether combo$Feature has been mapped back to the original ID.
  purrr::pmap_dfr(axes, function(Axis, ID_pattern, Name_pattern) {
    feats <- match_axis(feat_universe, ID_pattern, Name_pattern)
    if (!length(feats)) {
      return(tibble::tibble(Layer = layer, Axis = Axis,
                            Matched = 0L, Tested = 0L,
                            Up = 0L, Down = 0L, Sig = 0L,
                            `Med coef` = NA_real_))
    }
    feat_keys <- make.names(feats)
    sub <- combo |> dplyr::filter(make.names(Feature) %in% feat_keys)
    tibble::tibble(
      Layer      = layer,
      Axis       = Axis,
      Matched    = length(feats),
      Tested     = nrow(sub),
      Up         = sum(sub$Direction == "Up in HRSSL",  na.rm = TRUE),
      Down       = sum(sub$Direction == "Down in HRSSL", na.rm = TRUE),
      Sig        = sum(sub$sig_maaslin, na.rm = TRUE),
      `Med coef` = signif(median(sub$Coef, na.rm = TRUE), 2)
    )
  })
}

axis_path_tab <- axis_summary(combo_path, path_cpm_f, pathway_axes, "Pwy")
axis_ko_tab   <- axis_summary(combo_ko,   ko_cpm_f,   ko_axes,      "KO")

dplyr::bind_rows(axis_path_tab, axis_ko_tab) |>
  dplyr::mutate(`Med coef` = ifelse(is.na(`Med coef`), "", sprintf("%+.2f", `Med coef`))) |>
  kbl(booktabs = TRUE, longtable = TRUE,
      caption = "Directional summary across curated CRC / SSL functional axes. Matched: features in the layer matching the axis. Tested: features that passed prevalence filtering. Up/Down/Sig counted among Tested. Sig: features called by MaAsLin2 (q < 0.25). The rightmost column gives the median MaAsLin2 coefficient per axis as a direction indicator independent of significance.") |>
  kable_styling(latex_options = c("repeat_header", "hold_position"),
                font_size = 8) |>
  column_spec(2, width = "5.0cm")
```

```{r axis-plot-helper}
plot_axis_features <- function(combo, mat_cpm, axis_name, id_pat, name_pat,
                                group_vec, max_features = 12) {
  feats <- match_axis(rownames(mat_cpm), id_pat, name_pat)
  if (!length(feats)) {
    return(ggplot() +
             annotate("text", x = 0.5, y = 0.5,
                      label = paste(axis_name, "- no features matched"),
                      size = 4) +
             theme_void())
  }
  # Filter via make.names() so matching is robust to MaAsLin2's name
  # sanitisation regardless of how combo$Feature is stored.
  feat_keys <- make.names(feats)
  sub <- combo |>
    dplyr::filter(make.names(Feature) %in% feat_keys) |>
    dplyr::arrange(dplyr::coalesce(q_value, 1)) |>
    utils::head(max_features)
  if (!nrow(sub)) {
    return(ggplot() +
             annotate("text", x = 0.5, y = 0.5,
                      label = paste(axis_name, "- features matched but not tested"),
                      size = 4) +
             theme_void())
  }
  # sub$Feature can be either sanitised or original; resolve back to the
  # matrix rownames so subsetting works regardless.
  ord_keys   <- make.names(sub$Feature)
  mat_keys   <- make.names(rownames(mat_cpm))
  ord_rownames <- rownames(mat_cpm)[match(ord_keys, mat_keys)]
  ord_rownames <- ord_rownames[!is.na(ord_rownames)]
  # Wrap long pathway / KO names onto two lines so the strip text fits inside
  # the narrow 3-column facets (previously str_trunc(55) was producing
  # truncated single-line labels that got clipped by the panel width).
  wrap_strip <- function(x) stringr::str_wrap(x, width = 30)
  long <- as.data.frame(mat_cpm[ord_rownames, , drop = FALSE]) |>
    tibble::rownames_to_column("Feature") |>
    tidyr::pivot_longer(-Feature, names_to = "SampleID", values_to = "CPM") |>
    dplyr::left_join(data.frame(SampleID = colnames(mat_cpm), Group = group_vec),
                     by = "SampleID") |>
    dplyr::mutate(Feature = factor(wrap_strip(Feature),
                                   levels = wrap_strip(ord_rownames)),
                  CPM = CPM + 1)

  ggplot(long, aes(Group, CPM, fill = Group)) +
    geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.8) +
    geom_jitter(width = 0.12, size = 0.7, alpha = 0.8) +
    facet_wrap(~ Feature, scales = "free_y", ncol = 3) +
    scale_y_log10(labels = label_log()) +
    scale_fill_manual(values = group_colours, guide = "none") +
    labs(title = axis_name, x = NULL, y = "CPM + 1 (log scale)") +
    theme(strip.text = element_text(size = 6.5, lineheight = 0.9,
                                    margin = margin(2, 0, 2, 0)),
          axis.text.x = element_text(size = 8))
}

# Convenience: pull patterns for an axis by name from the curated tribbles
axis_pats <- function(axis_tbl, axis) {
  row <- axis_tbl[axis_tbl$Axis == axis, ]
  list(id = row$ID_pattern, name = row$Name_pattern)
}
```

## Butyrate / SCFA production (pathways)

```{r axis-butyrate-path, fig.height=5.2, fig.cap="Butyrate-associated MetaCyc pathways. Depletion in HRSSL would be consistent with the CRC literature."}
p <- axis_pats(pathway_axes, "Butyrate / SCFA production")
plot_axis_features(combo_path, path_cpm_f, "Butyrate / SCFA production",
                   p$id, p$name, meta$Group)
```

## Sulfur / hydrogen sulfide (pathways)

```{r axis-sulfur-path, fig.height=5.2, fig.cap="Sulfur metabolism and putative H2S generation pathways."}
p <- axis_pats(pathway_axes, "Sulfur / hydrogen sulfide")
plot_axis_features(combo_path, path_cpm_f, "Sulfur / hydrogen sulfide",
                   p$id, p$name, meta$Group)
```

## Polyamine biosynthesis (pathways)

```{r axis-polyamine-path, fig.height=5.2, fig.cap="Polyamine biosynthesis pathways - putrescine and spermidine routes are typically enriched in invasive CRC microbiomes."}
p <- axis_pats(pathway_axes, "Polyamine biosynthesis")
plot_axis_features(combo_path, path_cpm_f, "Polyamine biosynthesis",
                   p$id, p$name, meta$Group)
```

## Arginine / ornithine catabolism (pathways)

```{r axis-argorn-path, fig.height=5.2, fig.cap="Arginine and ornithine catabolic pathways. The cluster runs coherently down in HRSSL across all six matched pathways (none individually significant at q < 0.25), which is what makes it worth reading as an axis rather than feature by feature."}
p <- axis_pats(pathway_axes, "Arginine / ornithine catabolism")
plot_axis_features(combo_path, path_cpm_f, "Arginine / ornithine catabolism",
                   p$id, p$name, meta$Group)
```

## BCAA biosynthesis (pathways)

```{r axis-bcaa-path, fig.height=5.2, fig.cap="Branched-chain amino acid biosynthesis pathways. Most trend higher in HRSSL, though none reach q < 0.25 in this cohort."}
p <- axis_pats(pathway_axes, "BCAA biosynthesis (Val/Leu/Ile)")
plot_axis_features(combo_path, path_cpm_f, "BCAA biosynthesis (Val/Leu/Ile)",
                   p$id, p$name, meta$Group)
```

## LPS / KDO / ECA biosynthesis (pathways)

```{r axis-lps-path, fig.height=5.2, fig.cap="LPS, KDO and enterobacterial common antigen biosynthesis. Proxies for Proteobacteria abundance; ECASYN-PWY is markedly lower in HRSSL."}
p <- axis_pats(pathway_axes, "LPS / KDO / ECA (Proteobacteria)")
plot_axis_features(combo_path, path_cpm_f, "LPS / KDO / ECA (Proteobacteria)",
                   p$id, p$name, meta$Group)
```

## Folate / B-vitamin biosynthesis (pathways)

```{r axis-bvit-path, fig.height=5.2, fig.cap="Folate, B12, riboflavin, biotin, thiamine, pantothenate."}
# Combine folate + B12 + other B-vitamins for the figure
folate_feats <- c(match_axis(rownames(path_cpm_f),
                             pathway_axes$ID_pattern[pathway_axes$Axis == "Folate / one-carbon"],
                             pathway_axes$Name_pattern[pathway_axes$Axis == "Folate / one-carbon"]),
                  match_axis(rownames(path_cpm_f),
                             pathway_axes$ID_pattern[pathway_axes$Axis == "B12 / cobalamin"],
                             pathway_axes$Name_pattern[pathway_axes$Axis == "B12 / cobalamin"]),
                  match_axis(rownames(path_cpm_f),
                             pathway_axes$ID_pattern[pathway_axes$Axis == "Other B-vitamins (B1/B2/B5/B7)"],
                             pathway_axes$Name_pattern[pathway_axes$Axis == "Other B-vitamins (B1/B2/B5/B7)"]))
# Use an ID pattern that matches the union; name pattern empty to avoid double matching
id_union <- if (length(folate_feats)) paste0("^(", paste(folate_feats, collapse = "|"), ")$") else "^$"
plot_axis_features(combo_path, path_cpm_f, "Folate / B-vitamin biosynthesis",
                   id_union, "(?!.*)", meta$Group)
```

## Bile acid 7-alpha-dehydroxylation (KO panel)

```{r axis-bai-ko, fig.height=4.2, fig.cap="bai operon KOs - K15868-K15875 cover the 7-alpha-dehydroxylation cluster."}
p <- axis_pats(ko_axes, "Bile acid 7alpha-dehydroxylation")
plot_axis_features(combo_ko, ko_cpm_f, "Bile acid 7alpha-dehydroxylation",
                   p$id, p$name, meta$Group)
```

## Sulfite reduction / cysteine -> H2S (KO panel)

```{r axis-h2s-ko, fig.height=5, fig.cap="dsr operon, sulfate adenylyltransferase, and cysteine gamma-lyase paths to hydrogen sulfide."}
p_dsr  <- axis_pats(ko_axes, "Dissimilatory sulfite reductase (dsr)")
p_cys  <- axis_pats(ko_axes, "Cysteine -> H2S (gamma-lyase etc.)")
id_h2s   <- paste0("(", p_dsr$id,  "|", p_cys$id,  ")")
name_h2s <- paste0("(", p_dsr$name, "|", p_cys$name, ")")
plot_axis_features(combo_ko, ko_cpm_f, "Sulfite reduction / H2S",
                   id_h2s, name_h2s, meta$Group)
```

## Colibactin (pks island, clb) (KO)

```{r axis-clb-ko, fig.height=4.2, fig.cap="Colibactin biosynthesis (pks+ E. coli). Typically near-absent in stool metagenomes; absence is informative."}
p <- axis_pats(ko_axes, "Colibactin (clb / pks island)")
plot_axis_features(combo_ko, ko_cpm_f, "Colibactin (clb / pks island)",
                   p$id, p$name, meta$Group)
```

# Stratified contributions: who is driving the top pathways?

HUMAnN3 reports each pathway's abundance in two complementary ways. The
*unstratified* layer is the community-level total -- "sulfite reduction is at
500 CPM in this sample." The *stratified* layer breaks the same total down by
the species whose genes supplied the reads -- "of that 500 CPM, 410 came from
*Bilophila wadsworthia*, 70 from *Desulfovibrio piger*, the rest from
unclassified reads." All differential abundance testing in section 6 runs on
the unstratified layer, so a DA hit tells you that a function is shifted but
not who is shifting it. The stratified output is what closes that gap,
turning "sulfite reduction is enriched in HRSSL" into "*B. wadsworthia* is
expanded in HRSSL and is responsible for the sulfite reduction signal".

```{r stratified-prep}
# Stratified rows are of the form "PATHWAY|genus.species"
strat_df <- as.data.frame(path_strat) |>
  tibble::rownames_to_column("FullFeature") |>
  dplyr::filter(grepl("\\|", FullFeature)) |>
  tidyr::separate(FullFeature, into = c("Pathway", "Taxon"),
                  sep = "\\|", extra = "merge", fill = "right") |>
  tidyr::pivot_longer(-c(Pathway, Taxon), names_to = "SampleID", values_to = "CPM") |>
  dplyr::left_join(data.frame(SampleID = samples, Group = meta$Group),
                   by = "SampleID")
```

```{r stratified-top-pwys}
# Rank DA pathways by MaAsLin2 significance
da_pwys_ranked <- combo_path |>
  dplyr::filter(sig_maaslin) |>
  dplyr::arrange(dplyr::coalesce(q_value, 1)) |>
  dplyr::pull(Feature)

strat_pwys_available <- unique(strat_df$Pathway)

# Try top-6 DA pathways first; otherwise any DA pathway with stratified data;
# otherwise fall back to the 6 most-abundant pathways in the stratified output.
pwys_to_plot <- intersect(utils::head(da_pwys_ranked, 6), strat_pwys_available)
plot_source  <- "top-6 DA pathways (with stratified contributions)"

if (length(pwys_to_plot) == 0 && length(da_pwys_ranked) > 0) {
  pwys_to_plot <- utils::head(intersect(da_pwys_ranked, strat_pwys_available), 6)
  plot_source  <- "DA pathways with stratified contributions (top-6 had none)"
}

if (length(pwys_to_plot) == 0) {
  pwys_to_plot <- strat_df |>
    dplyr::group_by(Pathway) |>
    dplyr::summarise(total = sum(CPM, na.rm = TRUE), .groups = "drop") |>
    dplyr::arrange(dplyr::desc(total)) |>
    utils::head(6) |>
    dplyr::pull(Pathway)
  plot_source <- "top-6 most-abundant pathways (no DA pathway had stratified data)"
}

# Diagnostics
n_da              <- length(da_pwys_ranked)
n_da_with_strat   <- length(intersect(da_pwys_ranked, strat_pwys_available))
n_top6_with_strat <- length(intersect(utils::head(da_pwys_ranked, 6),
                                      strat_pwys_available))
```

The panels below show the top 6 DA pathways by significance with their
per-sample, per-species contributions (source: `r plot_source`). Of `r n_da`
DA pathways (MaAsLin2), `r n_da_with_strat` have stratified data
available and `r n_top6_with_strat` of the top 6 plotted here do. When fewer
than 6 do, the missing pathways fall into one of two cases:

* \textbf{community-level only} -- HUMAnN3 couldn't confidently attribute the
  underlying gene families to any single species, usually because the reads
  multi-mapped across taxa; or
* \textbf{attributed entirely to \texttt{unclassified} reads} -- the
  supporting reads didn't hit the species-level reference at all, so the
  function is real but the responsible taxon either isn't in the database or
  is too divergent to be assigned.

Both patterns are biologically informative rather than pipeline errors: a DA
pathway with no species attribution flags a functional signal whose
responsible taxon is generic, novel, or under-characterised -- exactly the
kind of finding that motivates follow-up work like MAG assembly or targeted
culture.

```{r stratified-top-plots, fig.height=6, fig.cap="Per-sample stacked contributions to the selected pathways. Each bar is one sample; colour blocks are contributing species (or 'unclassified' reads). Bars are grouped by HRSSL/NN status."}
strat_top <- strat_df |>
  dplyr::filter(Pathway %in% pwys_to_plot)

if (nrow(strat_top) == 0) {
  # Final guard - should be unreachable given the fallback above
  ggplot() +
    annotate("text", x = 0.5, y = 0.5,
             label = "No stratified pathway contributions available.") +
    theme_void()
} else {
  # HUMAnN3 stratified taxon strings have the form
  # "g__Genus.s__Genus_species" (or just "unclassified"). The full strings
  # don't fit four-across in the legend at page width, so trim to the species
  # binomial for display and drop the legend to three columns.
  clean_taxon <- function(x) {
    has_sp <- grepl("\\.s__[^.]+$", x)
    out    <- ifelse(has_sp, sub("^.*\\.s__", "", x), x)
    out    <- sub("^g__", "", out)
    gsub("_", " ", out)
  }

  strat_top <- strat_top |>
    dplyr::group_by(Pathway, Taxon) |>
    dplyr::mutate(total = sum(CPM, na.rm = TRUE)) |>
    dplyr::group_by(Pathway) |>
    dplyr::mutate(rk    = dplyr::dense_rank(-total),
                  Taxon = ifelse(rk <= 6, Taxon, "Other")) |>
    dplyr::ungroup() |>
    dplyr::group_by(Pathway, Taxon, SampleID, Group) |>
    dplyr::summarise(CPM = sum(CPM, na.rm = TRUE), .groups = "drop") |>
    dplyr::mutate(Pathway = stringr::str_wrap(Pathway, width = 45),
                  Taxon   = clean_taxon(Taxon))

  # Custom 17-colour categorical palette supplied for this plot. Real
  # species are assigned in alphabetical order from the start of the palette;
  # Other and unclassified are pushed to the end of the legend and pick up
  # the muted/grey colours at the tail of the palette. If a future run has
  # more than 17 unique taxa, the palette is interpolated to fit; if fewer,
  # the leading colours are used.
  custom_pal <- c("#ffd700", "#ffb14e", "#fa8775", "#f6327d", "#cd34b5",
                  "#2d0388", "#0000ff", "#a73e3e", "#83ff9d", "#918cba",
                  "#e7d0ff", "#00ffeb", "#64b06f", "#714848", "#000000",
                  "#447089", "#696969")

  taxa_present     <- unique(strat_top$Taxon)
  special_in_order <- c("Other", "unclassified")
  real_taxa        <- sort(setdiff(taxa_present, special_in_order))
  present_special  <- intersect(special_in_order, taxa_present)
  taxa_levels      <- c(real_taxa, present_special)
  n_taxa           <- length(taxa_levels)

  pal_full <- if (n_taxa <= length(custom_pal)) {
    custom_pal[seq_len(n_taxa)]
  } else {
    grDevices::colorRampPalette(custom_pal)(n_taxa)
  }
  # Force the last colours in the palette onto Other / unclassified so the
  # neutral greys at the tail of the supplied palette always land on the
  # catch-all categories, regardless of how many real taxa there are.
  if (length(present_special)) {
    n_special <- length(present_special)
    pal_full[(n_taxa - n_special + 1):n_taxa] <-
      utils::tail(custom_pal, n_special)
  }
  pal <- stats::setNames(pal_full, taxa_levels)

  strat_top$Taxon <- factor(strat_top$Taxon, levels = taxa_levels)

  ggplot(strat_top,
         aes(stats::reorder(SampleID, as.numeric(Group)), CPM, fill = Taxon)) +
    geom_col(width = 0.95) +
    facet_wrap(~ Pathway, scales = "free_y", ncol = 2) +
    scale_fill_manual(values = pal, drop = FALSE) +
    labs(x = NULL, y = "Contribution (CPM)", fill = NULL) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
          legend.position = "bottom",
          legend.text = element_text(size = 7),
          legend.key.size = unit(0.35, "cm"),
          strip.text = element_text(size = 7, lineheight = 0.9,
                                    margin = margin(2, 0, 2, 0))) +
    guides(fill = guide_legend(ncol = 3, byrow = TRUE))
}
```

# Biological interpretation

The targeted axes in section 8 are organised around the principal functional
shifts that recur in the CRC and serrated-neoplasia microbiome literature.
The directionality table in that section provides the per-axis signal even
where individual features fail FDR with n = 12 per group; interpret these as
hypothesis-generating priors rather than confirmatory findings.

Key axes and what to look for:

\begin{description}
  \item[\textbf{Butyrate biosynthesis}] Depletion in HRSSL would align with
        the well-documented loss of \textit{Faecalibacterium},
        \textit{Roseburia}, and other Lachnospiraceae butyrate producers
        along the serrated and adenoma-carcinoma pathways. Butyrate is the
        primary energy source for colonocytes and a HDAC inhibitor with
        anti-proliferative effects on transformed cells.
  \item[\textbf{Bile acid 7-alpha-dehydroxylation}] Enrichment of the
        \textit{bai} operon (KOs K15868-K15875), typically attributable to
        Clostridium cluster XIVa and \textit{C. scindens}, increases
        secondary bile acid (deoxycholate, lithocholate) production. These
        are colon-tumor-promoting in rodent models and elevated in human
        CRC stool.
  \item[\textbf{Hydrogen sulfide production}] The dissimilatory sulfite
        reductase pathway (K11180/K11181, dsrAB) and cysteine
        gamma-lyase routes (K01758) generate H2S, which is genotoxic to
        colonocytes at micromolar concentrations. \textit{Bilophila
        wadsworthia} and \textit{Desulfovibrio} species are the canonical
        contributors.
  \item[\textbf{Polyamine biosynthesis}] Putrescine and spermidine support
        rapid cell proliferation; microbial polyamine output complements
        host upregulation in neoplastic tissue. Ornithine decarboxylase
        (K01581) and related enzymes are the readout here.
  \item[\textbf{Mucin / glycan utilisation}] Sialidases, sulfatases, and
        fucosidases enable mucolytic species (e.g., \textit{Akkermansia
        muciniphila}, \textit{Ruminococcus gnavus}, \textit{R. torques}) to
        degrade the protective mucus layer. The directionality is
        context-dependent in serrated lesions; \textit{R. gnavus} expansion
        in particular has been associated with mucosal inflammation.
  \item[\textbf{Colibactin (pks island)}] The \textit{clb} cluster encodes
        colibactin, a genotoxic small molecule produced by pks+
        \textit{E. coli} that induces a specific mutational signature
        observed in human CRC genomes. Detection is sparse but biologically
        important.
  \item[\textbf{B-vitamin biosynthesis}] Folate biosynthesis (and
        consumption) by gut microbes intersects with host one-carbon
        metabolism. Cobalamin (B12) routes can serve as proxies for
        community structure shifts toward producer taxa.
  \item[\textbf{LPS / oxidative stress}] Enrichment is consistent with
        Proteobacteria expansion and mucosal inflammation in dysbiosis.
\end{description}

The MaAsLin2 list should be read alongside the cohort's compositional
structure. As discussed in section 6, HRSSL samples carry lower feature
richness than NN; a per-layer richness covariate has been entered into both
models to absorb the closure-driven component that CPM scaling would otherwise
leave in the `Group` effect. The resulting hit lists are therefore more
defensible than an unadjusted model would produce, though with n = 12 per
group, axis-level directionality (consistent direction across multiple related
features, section 8) remains an important complementary lens. A coherent shift
across an entire functional axis is less likely to reflect residual artefact
than any single feature's q-value, and the directional summary in section 8
provides that view, with per-axis Up/Down counts and median coefficient as the
axis-level signal.

# Session info

```{r sessioninfo, echo=FALSE}
sessionInfo()
```
